---
title: "Amazon Reviews Sentiment Analysis Lexicon Comparison"
author: "MRSadik"
date: "2026-03-29"

output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---




## Section 1: Load and Prepare Data
This section clears the environment, loads the dataset, and maps the 5-star rating system to binary sentiment labels, excluding neutral (3-star) reviews.


```{r step1, include=TRUE}
# Global options for the document
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10)


# Load required libraries
library(syuzhet)
library(sentimentr)
library(meanr)
library(paletteer)
library(dplyr)
library(caret)
library(ggplot2)
library(forcats)
library(mltools)

# Clear workspace
rm(list = ls())

# Load data using your specific Mac file path
data <- read.csv("/Users/rezwanesadik/Documents/DeskTop Old/Conference/AllProductReviews.csv", header = TRUE)    
# read.csv(file.choose(), header = TRUE)



# Prepare labels: 1,2 -> Negative; 4,5 -> Positive; 3 -> Excluded
data$label <- factor(data$ReviewStar, 
                     levels = c(1, 2, 4, 5), 
                     exclude = 3, 
                     labels = c("Negative", "Negative", "Positive", "Positive"))

# Remove empty reviews and NAs
data_amazon <- na.omit(data[data$ReviewBody != "" & !is.na(data$ReviewBody), ])

# Preview data
head(data_amazon)


```



## Section 2: Sentiment Analysis Scoring
We apply 11 different lexicon methods to calculate sentiment scores. These are then discretized into "Positive", "Negative", or "Neutral" based on their numerical value.

```{r step2, include=TRUE}

library(DT) # Added for interactive tables

# Global options for the document
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10)

afinn <- as_key(syuzhet:::afinn)

# Calculate sentiment scores (Lines broken down to prevent HTML overflow)
out1 <- data.frame(
  data_amazon,
  sentimentr_hu_liu = sentiment_by(
    data_amazon$ReviewBody, polarity_dt = lexicon::hash_sentiment_huliu, question.weight = 0)[["ave_sentiment"]],
  sentimentr_sentiword = sentiment_by(
    data_amazon$ReviewBody, polarity_dt = lexicon::hash_sentiment_sentiword, question.weight = 0)[["ave_sentiment"]],
  sentimentr_jockers = sentiment_by(
    data_amazon$ReviewBody, polarity_dt = lexicon::hash_sentiment_jockers, question.weight = 0)[["ave_sentiment"]],
  sentimentr_afinn = sentiment_by(
    data_amazon$ReviewBody, polarity_dt = afinn, question.weight = 0)[["ave_sentiment"]],
  sentimentr_nrc = sentiment_by(
    data_amazon$ReviewBody, polarity_dt = lexicon::hash_sentiment_nrc, question.weight = 0)[["ave_sentiment"]],
  sentimentr_loughran_mcdonald = sentiment_by(
    data_amazon$ReviewBody, polarity_dt = lexicon::hash_sentiment_loughran_mcdonald, question.weight = 0)[["ave_sentiment"]],
  meanr = meanr::score(data_amazon$ReviewBody)$score,
  syuzhet_ = setNames(as.data.frame(lapply(c("syuzhet", "bing", "afinn", "nrc"),
                                           function(x) get_sentiment(data_amazon$ReviewBody, method=x))),
                      paste0("", c("syuzhet", "bing", "afinn", "nrc"))),
  stringsAsFactors = FALSE
)

# Isolate the 11 lexicon score columns
lexicon_cols <- c("sentimentr_hu_liu", "sentimentr_sentiword", "sentimentr_jockers", 
                  "sentimentr_afinn", "sentimentr_nrc", "sentimentr_loughran_mcdonald", 
                  "meanr", "syuzhet_.syuzhet", "syuzhet_.bing", "syuzhet_.afinn", "syuzhet_.nrc")

out <- out1[, lexicon_cols]

# Map numerical scores to discrete labels
out[out > 0] <- "Positive"
out[out < 0] <- "Negative"
out[out == 0 ] <- "Neutral"

# Append text and true labels
out$text <- out1$ReviewBody
out$label <- out1$label

# Use DT for an interactive, searchable table instead of static kable
datatable(head(out[,1:5], 100), 
          options = list(pageLength = 5, scrollX = TRUE), 
          caption = "Sample of Categorized Sentiment Scores (Interactive)")

```



## Section 3: Accuracy Calculation
We compare the predicted labels from each lexicon against the original star-rating labels to determine which lexicon performs best.


```{r step3, include=TRUE}
# Global options for the document
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10)

reference <- out$label

# Define accuracy function
calculate_accuracy <- function(predicted, reference, lexicon_name) {
  u <- union(predicted, reference)
  t <- table(factor(predicted, u), factor(reference, u))
  accuracy <- tryCatch(confusionMatrix(t)$overall[1], error = function(e) NA)
  data.frame(Lexicon = lexicon_name, Accuracy = accuracy)
}

# Calculate accuracy for all lexicons
df_results <- data.frame(Lexicon = character(), Accuracy = numeric())
for (lexicon in lexicon_cols) {
  predicted <- out[[lexicon]]
  accuracy_df <- calculate_accuracy(predicted, reference, lexicon)
  df_results <- bind_rows(df_results, accuracy_df)
}

# Sort results
sorted_df <- df_results[order(df_results$Accuracy, decreasing = TRUE), ]

# Use DT to make the rankings interactive and format as percentages
datatable(sorted_df, 
          options = list(pageLength = 11, dom = 't'), 
          rownames = FALSE,
          caption = "Lexicon Accuracy Rankings") %>%
  formatPercentage('Accuracy', 1)




```




## Section 4: Accuracy Visualization
A comparison of the accuracy rates across all tested lexicons.

```{r step4, include=TRUE}
library(plotly) # Added for interactive plots

# Global options for the document
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10)

# Create the base ggplot
acc_plot <- ggplot(sorted_df, aes(x = fct_reorder(Lexicon, Accuracy), y = Accuracy, fill = Lexicon, text = paste("Accuracy:", sprintf("%.1f%%", Accuracy * 100)))) +
  scale_fill_brewer(palette = "Paired", type = "qual") + 
  geom_col() +
  labs(x = "Lexicon", y = "Accuracy", title = "Lexicon Based Sentiment Accuracy Comparison") +
  theme_minimal() + 
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1)) + 
  coord_flip()

# Convert ggplot to interactive plotly graph
ggplotly(acc_plot, tooltip = "text")

```

## Section 5: Valence Shifter Analysis
In this section, we analyze how valence shifters (negators, amplifiers, de-amplifiers, and adversative conjunctions) affect the sentiment of the Amazon reviews.

```{r step5, include=TRUE}
# Global options for the document
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10)

library(stringr)
library(lexicon)
library(knitr)
library(kableExtra)

# 1. Prepare text for manual extraction
text_clean <- tolower(as.character(out$text))
text_clean <- text_clean[!is.na(text_clean)]

# 2. Load valence shifter dictionary
shifters <- lexicon::hash_valence_shifters

# 3. Build exact word boundary regex patterns for each shifter type
regex_neg <- paste0("\\b(", paste(shifters$x[shifters$y == 1], collapse = "|"), ")\\b")
regex_amp <- paste0("\\b(", paste(shifters$x[shifters$y == 2], collapse = "|"), ")\\b")
regex_deamp <- paste0("\\b(", paste(shifters$x[shifters$y == 3], collapse = "|"), ")\\b")
regex_adv <- paste0("\\b(", paste(shifters$x[shifters$y == 4], collapse = "|"), ")\\b")

# 4. Count how many reviews contain at least one of each shifter type
count_neg <- sum(str_detect(text_clean, regex_neg))
count_amp <- sum(str_detect(text_clean, regex_amp))
count_deamp <- sum(str_detect(text_clean, regex_deamp))
count_adv <- sum(str_detect(text_clean, regex_adv))

total_reviews <- length(text_clean)

# 5. Calculate percentages relative to the total number of reviews
pct_neg <- paste0(round((count_neg / total_reviews) * 100), "%")
pct_amp <- paste0(round((count_amp / total_reviews) * 100), "%")
pct_deamp <- paste0(round((count_deamp / total_reviews) * 100), "%")
pct_adv <- paste0(round((count_adv / total_reviews) * 100), "%")

# 6. Construct the formatted dataframe
valance_table <- data.frame(
  Text = "Amazon_Reviews",
  Negator = pct_neg,
  Amplifier = pct_amp,
  Deamplifier = pct_deamp,
  Adversative = pct_adv,
  stringsAsFactors = FALSE
)


# 7. Render the styled table to match specifications
kable(valance_table, align = "c", caption = "Table 5: Valance Shifters output of the dataset.") %>%
  kable_styling(bootstrap_options = "bordered", full_width = FALSE, position = "center", font_size = 14) %>%
  row_spec(0, bold = TRUE, font_size = 14) %>%
  column_spec(1, bold = TRUE, italic = TRUE) %>%
  column_spec(2, background = "#FCE4D6", bold = TRUE) %>%
  column_spec(3, background = "#E2EFDA", bold = TRUE) %>%
  column_spec(4, background = "#F2F2F2", bold = TRUE) %>%
  column_spec(5, background = "#DDEBF7", bold = TRUE)





```

## Section 6: Error Analysis and Confusion Matrix
We isolate the predictions of the Jockers lexicon, filtering out common ambiguous (Neutral) predictions to assess its binary classification capabilities via a confusion matrix and Matthews correlation coefficient (MCC).


```{r step6, include=TRUE}
# Filter for reviews predicted as Neutral by all key lexicons
common_neutral_reviews <- out %>%
  filter(
    sentimentr_jockers == "Neutral" & syuzhet_.afinn == "Neutral" &
    sentimentr_hu_liu == "Neutral" & sentimentr_sentiword == "Neutral" &
    sentimentr_loughran_mcdonald == "Neutral" & sentimentr_afinn == "Neutral" &
    sentimentr_nrc == "Neutral" & meanr == "Neutral" &
    syuzhet_.syuzhet == "Neutral" & syuzhet_.bing == "Neutral" &
    syuzhet_.nrc == "Neutral"
  )

# Remove the common neutral reviews for binary evaluation
final_out <- out[!(out$text %in% common_neutral_reviews$text), ]

predicted_j <- factor(final_out$sentimentr_jockers, levels = c("Negative", "Positive"))
reference_j <- factor(final_out$label, levels = c("Negative", "Positive"))

# Create 2x2 confusion matrix
conf_matrix_ <- table(predicted_j, reference_j)

# Visualize confusion matrix
fourfoldplot(conf_matrix_, color = c("pink", "lightblue"), conf.level = 0, margin = 1, main = "Jockers Lexicon Confusion Matrix")

# Calculate Matthews Correlation Coefficient (MCC)
mcc_value <- mcc(TN=conf_matrix_[1,1], TP=conf_matrix_[2,2], FP=conf_matrix_[1,2], FN=conf_matrix_[2,1])
cat("Matthews Correlation Coefficient (MCC):", round(mcc_value, 4), "\n\n")

# Detailed caret statistics
confusionMatrix(predicted_j, reference_j)

```